The evolution of gene expression in the term placenta of viviparous mammals

### 2015-10-08 manuscript/manuscript.Rnw
update the bonobo description and homo sapiens description

The top 50 expressed genes as measured by \ac{FPKM} in the term placenta of
the human are shown in \fref{tab:top50_human_placenta} with a list of all genes in
Supplemental Table X. The gene with by far the highest \ac{FPKM} value
(101962) in humans is chorionic somatomammotropin hormone 1 (CSH1),
followed by chorionic somatomammotropin hormone 2 (CSH2), and
chorionic somatomammotropin hormone-like 1 (CSHL1), all three of which
are members of a distinct expansion of the growth hormone gene family
within hominoids. Another member of the growth hormone gene family,
GH2 (\ac{FPKM}=729.639), is also expressed to a high degree in the human
placenta (Table 2). Ten pregnancy-specific beta-1-glycoprotein genes
are highly-expressed (FPKM>100) in the term placenta of humans, with
PSG4 showing the greatest degree of expression (\ac{FPKM}= 5201.31; Table
2). These genes are members of a hominoid-specific expansion of the
carcinoembryonic antigen (CEA) family and known to be expressed by the
syncytiotrophoblast (Lei et al., 1992; Rand and Cross, 2008). Genes
involved in the production of estrogen (CYP19A1) and progesterone
(HSD3B1) show exceptionally high expression in the human placenta
(\ac{FPKM} = 4817.15 and 829.846 respectively), as well as CGA (\ac{FPKM}=
4793.08), which generates the alpha subnit of multiple human chorionic
gonadotropin hormones (Table 2). The genes producing the beta subunits
of the chorionic gonadotropins (CGB, CGB1, CGB2, CGB5, CGB7, CGB8) are
either not expressed (CGB8) or expressed modestly (FPKM≤20.826; Table
8). Other highly expressed genes in the top 50 include members of the
hemoglobin family (e.g. HBA1, HBB, HBG2, HBA2), the maternally
imprinted long noncoding RNA gene H19, genes coding for proteins
required for growth and/or cell proliferation (ADAM12, KISS1, IGF2,
GPC3, CSF3R), anticoagulation (TFPI2), and glucose transport (SLC2A1)
among others (Table 2, Supplemental Table X)

@@

The top 50 expressed genes as measured by \ac{FPKM} in the term placenta of
the bonobo are shown in Table 3 with a list of all genes in
Supplemental Table X. Expression in the bonobo placenta differs
somewhat from that of the human, as evidenced by the makeup of the top
50 genes (Table 3). The highest expression (\ac{FPKM}=10149.1) in bonobo is
proteoglycan 2, bone marrow (PRG2), which is involved in insulin
growth factor signaling and angiogenesis during pregnancy (Weyer and
Glerup, 2011). PRG2 is also expressed at relatively high levels in the
human placenta (\ac{FPKM}=250.675; Supplemental Table X), although at
comparatively much lower levels than the bonobo. A bonobo gene closely
related to the chorionic somatomammotropins (Q95MK7\_PANTR) is also
expressed at high levels (\ac{FPKM}=1232.56; Table 3). Two other Pan genes
included within this family (ENSPTRG00000030805 which shows similarity
to GH2 and ENSPTRG00000034560) show expression levels of \ac{FPKM}>89. Six
pregnancy-specific beta-1-glycoprotein genes were expressed in the
bonobo placenta with \ac{FPKM} ranging from 20.885-167.961 (Supplemental
Table X). Placental expression of the genes involved in estrogen and
progesterone production (CYP19A1, HSD3B1 respectively) show lower
expression in the bonobo as compared to humans (CYP19A1 \ac{FPKM}= 48.6877;
HSD3B1 \ac{FPKM}= 54.5301; Table 8). In addition, CGA is expressed in the
bonobo placenta (\ac{FPKM}=202.629), but at much lower levels than humans
as well (Table 8). Two Ensembl-annotated beta subunits of the
chorionic gonadotropin genes are expressed in bonobo, one related to
CGB and another related to CGB7; both are not highly expressed (Table
8). Among the other top 50 expressed genes, there is high expression
of ribosomal proteins (e.g., RPL18, RPL23), keratins (e.g., KRT7,
KRT19), heat-shock proteins (e.g., HSPA5, HSPA1A), serpins (SERPINE1,
SERPINE2), insulin growth factor 2 (IGF2) and the hormone
adrenomedullin (ADM).

>>>>

The top 50 expressed genes as measured by \ac{FPKM} in the term
placenta of \textit{Homo sapiens} are shown in
\fref{tab:top50_human_placenta} with a list of all genes in
Supplemental Table X. The gene with by far the highest \ac{FPKM} value
(\Sexpr{gene.expression(species.name="homo sapiens",gene="CSH1")}) %
in humans is \ac{CSH1}, followed by \ac{CSH2}, and \ac{CSHL1}, all
three of which are members of a distinct expansion of the growth
hormone gene family within hominoids. Another member of the growth
hormone gene family, \ac{GH2} (\ac{FPKM}=%
\Sexpr{gene.expression(species.name="homo sapiens",gene="GH2")}), %
is also expressed to a high degree in the human placenta
(\fref{tab:top50_human_placenta}).
% 
Ten pregnancy-specific beta-1-glycoprotein genes are highly-expressed
(\ac{FPKM}>100) in the term placenta of humans, with \ac{PSG4} showing the
greatest degree of expression %
(\ac{FPKM}=\Sexpr{gene.expression(species.name="homo sapiens",gene="PSG4")}). % 
These
genes are members of a hominoid-specific expansion of the
carcinoembryonic antigen (CEA) family and known to be expressed by the
syncytiotrophoblast (Lei et al., 1992; Rand and Cross, 2008). Genes
involved in the production of estrogen (CYP19A1) and progesterone
(HSD3B1) show exceptionally high expression in the human placenta
(\ac{FPKM} = 4817.15 and 829.846 respectively), as well as CGA
(\ac{FPKM}= 4793.08), which generates the alpha subnit of multiple
human chorionic gonadotropin hormones (Table 2). The genes producing
the beta subunits of the chorionic gonadotropins (CGB, CGB1, CGB2,
CGB5, CGB7, CGB8) are either not expressed (CGB8) or expressed
modestly (FPKM≤20.826; Table 8). Other highly expressed genes in the
top 50 include members of the hemoglobin family (e.g. HBA1, HBB, HBG2,
HBA2), the maternally imprinted long noncoding RNA gene H19, genes
coding for proteins required for growth and/or cell proliferation
(ADAM12, KISS1, IGF2, GPC3, CSF3R), anticoagulation (TFPI2), and
glucose transport (SLC2A1) among others (Table 2, Supplemental Table
X)

@@

The top 50 expressed genes as measured by \ac{FPKM} in the term placenta of
the bonobo are shown in \fref{tab:top50_bonobo_placenta} with a list of all genes in
Supplemental Table X. Expression in the bonobo placenta differs
somewhat from that of the human, as evidenced by the makeup of the top
50 genes (Table 3). The highest expression (\ac{FPKM}=%
\Sexpr{gene.expression("pan paniscus","PRG2")}%
%
) in bonobo is \ac{PRG2}, which is involved in insulin
growth factor signaling and angiogenesis during pregnancy \cite{Weyer.ea2011:Placentalregulationofpeptide}. \ac{PRG2} is also expressed at relatively high levels in the
human placenta (\ac{FPKM}=\Sexpr{gene.expression("homo sapiens","PRG2")}; Supplemental Table X), although at
comparatively much lower levels than the bonobo. A bonobo gene closely
related to the chorionic somatomammotropins (Q95MK7\_PANTR) is also
expressed at high levels (\ac{FPKM}=1232.56; \fref{tab:top50_bonobo_placenta}). Two other Pan genes
included within this family (ENSPTRG00000030805 which shows similarity
to GH2 and ENSPTRG00000034560) show expression levels of \ac{FPKM}>89. Six
pregnancy-specific beta-1-glycoprotein genes were expressed in the
bonobo placenta with \ac{FPKM} ranging from 20.885-167.961 (Supplemental
Table X). Placental expression of the genes involved in estrogen and
progesterone production (CYP19A1, HSD3B1 respectively) show lower
expression in the bonobo as compared to humans (CYP19A1 \ac{FPKM}= 48.6877;
HSD3B1 \ac{FPKM}= 54.5301; Table 8). In addition, CGA is expressed in the
bonobo placenta (\ac{FPKM}=202.629), but at much lower levels than humans
as well (Table 8). Two Ensembl-annotated beta subunits of the
chorionic gonadotropin genes are expressed in bonobo, one related to
CGB and another related to CGB7; both are not highly expressed (Table
8). Among the other top 50 expressed genes, there is high expression
of ribosomal proteins (e.g., RPL18, RPL23), keratins (e.g., KRT7,
KRT19), heat-shock proteins (e.g., HSPA5, HSPA1A), serpins (SERPINE1,
SERPINE2), insulin growth factor 2 (IGF2) and the hormone
adrenomedullin (ADM).
